Sparse Matrix Vector Multiplication with Multiple vectors and Multiple Matrices on the MIC Architecture

### 2013-10-18 ieee_paper.tex
GB/s -> ~GB/s; Gflop/s => ~Gflop/s

In Section~\ref{sec:rbf} we present the computation of RBFs and how it
can be expressed 16 multiplication of 4 vectors by 4 sparse matrices
with a common sparsity pattern. Section~\ref{sec:model} presents an
estimation of the instruction intensity of various form of the
computations and we show that a 6- fold improvement can be
expected when computing the 16 multiplications simultaneously and
reach a total of about 210 Gflop/s. This performance represents
approximatively 10\% of the available flop/s of a Knight Corner
coprocessor. Therefore, it is necessary to have implementations that
perform the computation in as little amount of instructions as
possible. We describe in Section~\ref{sec:impl} the details of the MIC
architecture and how to use specialized load, store, swizzle and
permutation instruction to efficiently bring the data in the vector
registers to be processed. Section~\ref{sec:expe} gives some
experimental result about the amount of bandwidth that can be achieved
depending on how the SpMV kernel is written and the actual performance
of the various kernel on multiple classes of matrices some generated
for analysis purpose and some extracted from and RBF application. A
performance of 135 Gflop/s is achieved on real scenario. Concluding
remarks and perspectives are provided in Section~\ref{sec:ccl}.

@@

Figure~\ref{fig:perf_predict} shows how these ratios translate
into actual performance. This figure provides projected best and worst
Gflop/s achievable assuming the computation is memory-bound and the
architecture reaches 150 Gflop/s. Notice that \cite{Saule13-ARXIV}
showed a higher peak bandwidth, but we will show in
Section~\ref{sec:expe} why 150Gflop/s is a better estimate of what one
might achieve in these kernels. In single precision, the worse that
one could achieve by using 4 vectors and 4 matrices is much higher
than the best achievable using a classical SpMV computation. The best
performance reachable is 210Gflop/s: almost six times higher than the
peak of the classical SpMV case.

\begin{figure}
  \centering 
  \includegraphics[width=.9\linewidth]{figures/gflops_peak.pdf}\label{fig:gflops-peak-perf}
  \caption{Estimation of the maximum achievable performance using a
    device with a memory to computational units of 150GB/s when
    varying the number of vectors and matrices.}
  \label{fig:perf_predict}
\end{figure}

@@

In this work, we use an Intel Xeon Phi 5110P coprocessor. This card
has 8 memory controllers that can each execute 5 billion
transactions per second. Each has two 32-bit channels, achieving a total
bandwidth of 320GB/s aggregated across all the memory
controllers. There are 60 cores clocked at 1.05GHz. Their memory
interfaces are 32-bit wide with two channels and the total bandwidth
is 8.4GB/s per core. Thus, the cores should be able to consume 504GB/s
at most. However, the bandwidth between the cores and the memory
controllers is limited by the ring network that connect the cores and
the memory controller. Its precise bandwidth is unknown but it is
believed to be between 200GB/s and 250GB/s.

@@

As explained in Section~\ref{sec:model}, we expect to achieve a performance of
about 210Gflop/s in single precision in the best case. We will focus
on the single precision case with four vectors and four matrices. 
Similar techniques apply for other matrix/vector combinations. This target
performance represents 10\% of the peak performance of the
architecture which is impossible without an efficient vectorization of
the kernel. Indeed, to reach such a performance, a Fused Multiply-Add
instruction on fully loaded registers must be executed at most every
10 cycles.

@@

When using the MIC card (Figure~\ref{fig:band_rw_MIC}) the {\tt
  read\_cpp} implementation reaches a performance of 160GB/s
\todo{gordon check exact number} which is better than the {\tt read}
implementation which only achieves 145GB/s \todo{and that one}. This
is explained by the compiler adding loop unrolling and software
prefetching in the assembly code in the {\tt read\_cpp} case which is
not added in the {\tt read} case. On the other hand, the {\tt write}
case reaches a performance of 155GB/s\todo{and that one} while the
{\tt write\_cpp} implementation never passes 55GB/s. This difference
is due to the use of the {\tt \_mm512\_store\_nrngo\_ps} instruction
which bypasses the Read For Ownership protocol and allows the writes
to be performed in any order. The compiler can not use this
instruction on its own because using would break the memory convention
useful for some parallel algorithms. The {\tt read/write} benchmark
performs at 140GB/s while the {\tt read/write cpp} benchmark performs
at about 90GB/s.

Note that SpMV kernels will likely use only simple read and simple
writes on the MIC architecture. We also benchmark the performance we
can achieve using {\tt gather} and {\tt unpack} operations and we
present the results in Figure~\ref{fig:band_gather}. The {\tt unpack}
instruction can reach a performance of 140GB/s. (There is no easy way
to implement the equivalent of the {\tt unpack} operation without
using explicit registers.) The performance of the {\tt gather}
operations depends in which order the array is read. If the indices
are ordered ({\tt compact}) then a performance of 140GB/s is
reached. If the indices are completely randomized, the performance
drops to 2GB/s\todo{correct?}. We can see that implementing the
indirection mechanism without vector instructions makes the
performance drop significantly. (In the {\tt compact} case, it never
passes 100GB/s and is close to 50GB/s most of the time.)

@@

To summarize, using vector instruction seems mandatory to reach the
highest bandwidth on the MIC architecture. For the instructions used
in a typical SpMV computation, one can expect a bandwidth between
140GB/s to 160GB/s. That is why we set in our estimation the best
achievable bandwidth to 150GB/s despite carefully crafted code to
maximize bandwidth can reach higher
values~\cite{Saule13-ARXIV}.

@@

We now investigate the actual performance that we can obtain when
multiplying 4 vectors by 4 matrices in single
precision. Figure~\ref{fig:expe_types} present results for different
types of matrices and also shows the minimum and maximum performance
predicted in Section~\ref{sec:model}. Figure~\ref{fig:perf_ideal}
gives the results for the {\tt supercompact} and {\tt random} matrices
which represent both the best and the worst case for such a
computation. The {\tt supercompact} case peaks at
205Gflop/s\todo{check that} being very close to the predicted peak
performance of 213Gflop/s. Conversely the performance of the {\tt
  random} case decreases to 59Gflop/s\todo{check} which is close to
the lowest predicted performance of 55Gflop/s. One can see that RCM
ordering helps the {\tt random} case but the impact decreases when the
size of the matrix increase.

Figure~\ref{fig:perf_realistic} gives the results on the {\tt compact}
case which represents the best realistic matrix one could find and in
an {\tt rbf} matrix extracted from a real application. In the {\tt
  compact} case, the performance can be as high as 195Gflop/s being
within 15\% of the peak predicted performance. The performance of the
{\tt rbf} case varies between 100Gflop/s and 140Gflop/s. Most of the
time, the RCM ordering provide an improvement which can be as high as
30Gflop/s.

@@

We finally investigate the central questions of this paper. Do we gain
actual performance by transforming classical SpMV computation into the
multiplication of 4 vectors by 4 matrices for the computation of the
derivative of RBFs? Does using manual vectorization improve actual
application performance? Figure~\ref{fig:comparison_spmv} compares the
performance achieved by a classical SpMV and by the multiplication of
4 vectors by 4 matrices using either standard C++ code and our
optimized implementation. The
classical SpMV computation reaches 15Gflop/s\todo{check}. The standard
C++ implementation reaches a performance of 40Gflop/s while our
optimized implementation reaches 137Gflop/s\todo{check}. Notice that
RCM ordering has no impact in the standard C++ implementation while it
provides an significant improvement in our manually vectorized
version. This highlights the C++ implementation is instruction bound
while our manually vectorized implementation is most likely memory
bound.

To put things in perspective, we conducted some experiments on a
classical CPU architecture with 2 Intel Core i7 processor, each has 8
cores which support AVX instructions. Both the compiler generated and
the manually written vectorization using AVX instruction reach a
(read) bandwidth of 70GB/s. This indicates that the performance of the
classical SpMV can not overcome 16.8Gflop/s. A generic implementation
of the multiplication of 4 vectors with 4 matrices on the {\tt RBF}
matrix reaches a performance of 27 Gflop/s; while a manual
implementation using AVX instruction reaches a performance of 42
Gflop/s. This confirms that our method is sound and generalizes to
other architecture than the Intel MIC architecture.

@@

Our implementation makes use of the IMCI MIC instruction set and
includes a number of swizzling and channel swapping operations, for an
extremely efficient tensor product implementation. On a $96^3$ 3D
grid, it reaches a performance of 140 Gflop/s which is 2.7 times
faster than the best possible performance achievable for a single
vector and a single derivative and more than 7 times faster than a
practical implementation.

%% Using a matrix with $96^3$ rows, we achieve a speedup of 7.7 relative
%% to the base case of one derivative and one function (1/1 case). We
%% have achieved a speed between 100 Gflop/s to 140 Gflop/s on 60 cores
%% using 240 threads, or a speed of 47\% to 66\% of the maximum possible
%% performance of this architecture. As a point of comparison, the
%% maximum possible speed for the 1/1 case is around 36 Gflop/s, of which
%% we achieve 18 Gflop/s. Our optimal implementation makes use of the
%% IMCI MIC instruction set which includes a number of swizzling and
%% channel swapping operations, for an extremely efficient tensor product
%% implementation. A straightforward implementation without vectorization
%% runs at around 35 Gflops, or 16\% of peak performance.

>>>>

In Section~\ref{sec:rbf} we present the computation of RBFs and how it
can be expressed 16 multiplication of 4 vectors by 4 sparse matrices
with a common sparsity pattern. Section~\ref{sec:model} presents an
estimation of the instruction intensity of various form of the
computations and we show that a 6- fold improvement can be
expected when computing the 16 multiplications simultaneously and
reach a total of about 210~Gflop/s. This performance represents
approximatively 10\% of the available flop/s of a Knight Corner
coprocessor. Therefore, it is necessary to have implementations that
perform the computation in as little amount of instructions as
possible. We describe in Section~\ref{sec:impl} the details of the MIC
architecture and how to use specialized load, store, swizzle and
permutation instruction to efficiently bring the data in the vector
registers to be processed. Section~\ref{sec:expe} gives some
experimental result about the amount of bandwidth that can be achieved
depending on how the SpMV kernel is written and the actual performance
of the various kernel on multiple classes of matrices some generated
for analysis purpose and some extracted from and RBF application. A
performance of 135~Gflop/s is achieved on real scenario. Concluding
remarks and perspectives are provided in Section~\ref{sec:ccl}.

@@

Figure~\ref{fig:perf_predict} shows how these ratios translate
into actual performance. This figure provides projected best and worst
Gflop/s achievable assuming the computation is memory-bound and the
architecture reaches 150~Gflop/s. Notice that \cite{Saule13-ARXIV}
showed a higher peak bandwidth, but we will show in
Section~\ref{sec:expe} why 150~Gflop/s is a better estimate of what one
might achieve in these kernels. In single precision, the worse that
one could achieve by using 4 vectors and 4 matrices is much higher
than the best achievable using a classical SpMV computation. The best
performance reachable is 210~Gflop/s: almost six times higher than the
peak of the classical SpMV case.

\begin{figure}
  \centering 
  \includegraphics[width=.9\linewidth]{figures/gflops_peak.pdf}\label{fig:gflops-peak-perf}
  \caption{Estimation of the maximum achievable performance using a
    device with a memory to computational units of 150~GB/s when
    varying the number of vectors and matrices.}
  \label{fig:perf_predict}
\end{figure}

@@

In this work, we use an Intel Xeon Phi 5110P coprocessor. This card
has 8 memory controllers that can each execute 5 billion
transactions per second. Each has two 32-bit channels, achieving a total
bandwidth of 320~GB/s aggregated across all the memory
controllers. There are 60 cores clocked at 1.05GHz. Their memory
interfaces are 32-bit wide with two channels and the total bandwidth
is 8.4~GB/s per core. Thus, the cores should be able to consume 504~GB/s
at most. However, the bandwidth between the cores and the memory
controllers is limited by the ring network that connect the cores and
the memory controller. Its precise bandwidth is unknown but it is
believed to be between 200~GB/s and 250~GB/s.

@@

As explained in Section~\ref{sec:model}, we expect to achieve a performance of
about 210~Gflop/s in single precision in the best case. We will focus
on the single precision case with four vectors and four matrices. 
Similar techniques apply for other matrix/vector combinations. This target
performance represents 10\% of the peak performance of the
architecture which is impossible without an efficient vectorization of
the kernel. Indeed, to reach such a performance, a Fused Multiply-Add
instruction on fully loaded registers must be executed at most every
10 cycles.

@@

When using the MIC card (Figure~\ref{fig:band_rw_MIC}) the {\tt
  read\_cpp} implementation reaches a performance of 160~GB/s
\todo{gordon check exact number} which is better than the {\tt read}
implementation which only achieves 145~GB/s \todo{and that one}. This
is explained by the compiler adding loop unrolling and software
prefetching in the assembly code in the {\tt read\_cpp} case which is
not added in the {\tt read} case. On the other hand, the {\tt write}
case reaches a performance of 155~GB/s\todo{and that one} while the
{\tt write\_cpp} implementation never passes 55~GB/s. This difference
is due to the use of the {\tt \_mm512\_store\_nrngo\_ps} instruction
which bypasses the Read For Ownership protocol and allows the writes
to be performed in any order. The compiler can not use this
instruction on its own because using would break the memory convention
useful for some parallel algorithms. The {\tt read/write} benchmark
performs at 140~GB/s while the {\tt read/write cpp} benchmark performs
at about 90~GB/s.

Note that SpMV kernels will likely use only simple read and simple
writes on the MIC architecture. We also benchmark the performance we
can achieve using {\tt gather} and {\tt unpack} operations and we
present the results in Figure~\ref{fig:band_gather}. The {\tt unpack}
instruction can reach a performance of 140~GB/s. (There is no easy way
to implement the equivalent of the {\tt unpack} operation without
using explicit registers.) The performance of the {\tt gather}
operations depends in which order the array is read. If the indices
are ordered ({\tt compact}) then a performance of 140~GB/s is
reached. If the indices are completely randomized, the performance
drops to 2~GB/s\todo{correct?}. We can see that implementing the
indirection mechanism without vector instructions makes the
performance drop significantly. (In the {\tt compact} case, it never
passes 100~GB/s and is close to 50~GB/s most of the time.)

@@

To summarize, using vector instruction seems mandatory to reach the
highest bandwidth on the MIC architecture. For the instructions used
in a typical SpMV computation, one can expect a bandwidth between
140~GB/s to 160~GB/s. That is why we set in our estimation the best
achievable bandwidth to 150~GB/s despite carefully crafted code to
maximize bandwidth can reach higher
values~\cite{Saule13-ARXIV}.

@@

We now investigate the actual performance that we can obtain when
multiplying 4 vectors by 4 matrices in single
precision. Figure~\ref{fig:expe_types} present results for different
types of matrices and also shows the minimum and maximum performance
predicted in Section~\ref{sec:model}. Figure~\ref{fig:perf_ideal}
gives the results for the {\tt supercompact} and {\tt random} matrices
which represent both the best and the worst case for such a
computation. The {\tt supercompact} case peaks at
205~Gflop/s\todo{check that} being very close to the predicted peak
performance of 213~Gflop/s. Conversely the performance of the {\tt
  random} case decreases to 59~Gflop/s\todo{check} which is close to
the lowest predicted performance of 55~Gflop/s. One can see that RCM
ordering helps the {\tt random} case but the impact decreases when the
size of the matrix increase.

Figure~\ref{fig:perf_realistic} gives the results on the {\tt compact}
case which represents the best realistic matrix one could find and in
an {\tt rbf} matrix extracted from a real application. In the {\tt
  compact} case, the performance can be as high as 195~Gflop/s being
within 15\% of the peak predicted performance. The performance of the
{\tt rbf} case varies between 100~Gflop/s and 140~Gflop/s. Most of the
time, the RCM ordering provide an improvement which can be as high as
30~Gflop/s.

@@

We finally investigate the central questions of this paper. Do we gain
actual performance by transforming classical SpMV computation into the
multiplication of 4 vectors by 4 matrices for the computation of the
derivative of RBFs? Does using manual vectorization improve actual
application performance? Figure~\ref{fig:comparison_spmv} compares the
performance achieved by a classical SpMV and by the multiplication of
4 vectors by 4 matrices using either standard C++ code and our
optimized implementation. The
classical SpMV computation reaches 15~Gflop/s\todo{check}. The standard
C++ implementation reaches a performance of 40~Gflop/s while our
optimized implementation reaches 137~Gflop/s\todo{check}. Notice that
RCM ordering has no impact in the standard C++ implementation while it
provides an significant improvement in our manually vectorized
version. This highlights the C++ implementation is instruction bound
while our manually vectorized implementation is most likely memory
bound.

To put things in perspective, we conducted some experiments on a
classical CPU architecture with 2 Intel Core i7 processor, each has 8
cores which support AVX instructions. Both the compiler generated and
the manually written vectorization using AVX instruction reach a
(read) bandwidth of 70~GB/s. This indicates that the performance of
the classical SpMV can not overcome 16.8~Gflop/s. A generic
implementation of the multiplication of 4 vectors with 4 matrices on
the {\tt RBF} matrix reaches a performance of 27~Gflop/s; while a
manual implementation using AVX instruction reaches a performance of
42~Gflop/s. This confirms that our method is sound and generalizes to
other architecture than the Intel MIC architecture.

@@

Our implementation makes use of the IMCI MIC instruction set and
includes a number of swizzling and channel swapping operations, for an
extremely efficient tensor product implementation. On a $96^3$ 3D
grid, it reaches a performance of 140~Gflop/s which is 2.7 times
faster than the best possible performance achievable for a single
vector and a single derivative and more than 7 times faster than a
practical implementation.
